MLH1 Results Outline

### 2020-02-28 doc/Results.rmd
cleaning up doc at the lib

<In order to deconstruct the cell wide MLH1 counts we look at the recombination landscape at the single bivalent level.>
<FROM number of crossovers per chromsoomes TO rec lanscape across single chromsomes>
<A big motivation was to quantifiy recombiantion landscape differences across single bivalents>

@@

@@

An additional measure of single bivalents, we tested a measure of the SC lengths for the whole genome (whole cells) using a image nalysis algorithm which measures the skeletonized bivalent and provides an acurate measure of the lengths of single SC/bivalents (ref RWang).

We apply a skeletonizing image analysis algorithm to all MLH1 meiocyte spreads. X cell wide total SC measures remained after and after outliers removed,
Mouse means were calculated for their total SC area per cell and are displayed in figure X.

@@

<We estimate the effect of the sex chromosomes from hand measured data...the XX (3rd longest chromsomes) in females is ~10% of total the summed SC area, the average propotion of the XY in spermatocytes is XX%?  however the morphology of the XY is highly variable across cells>

@@

In the mixed models for the sex differences mean **short bivalents** length as the dependant variable, sex is the most significant effect (LTR; p =`r Q1lmer.MM_SC_short_results$sex.results`). The interaction (subspecies by sex) effect was also slightly significant (p = `r Q1lmer.MM_SC_short_results$interaction.pval`).

For the mixed model run on the mean total SC metric, there are more significant effects for strain and subsp for the mixed model using mean **total sc** per cell as the dependant variable, the three fixed effects and one random effect were significant (LTR; subsp p= `r Q1.lmer.totSC_results$sub.sp`, sex p = `r Q1.lmer.totSC_results$sex.results`, and interaction p = `r Q1.lmer.totSC_results$interaction.pval`, (random LRT; strain p = `r Q1.lmer.totSC_results$rand.pval`).

@@

In linear models (M2 and M3) for **mean short bivalent length**, male is most significant (M2 glm; p =`r summ.Q1.glm_short.SC_M2$coefficients[,4][8]` and M3 glm; p = `r summ.Q1.glm_short.SC_M3$coefficients[,4][8]`). In the linear model M2 the skive strain and strain by sex effects are slightly significant (glm; p=`r summ.Q1.glm_short.SC_M2$coefficients[,4][7]`, p =`r summ.Q1.glm_short.SC_M2$coefficients[,4][14]` respectively), however these could be due to low sampling of skive females.

@@

<1. the longer female sc's is consistant and seen across all strains, even in the high rec strains, suggest a decoupling of sc and gwRR - when comparing across sexes -- or the sc area is used differently for CO in oocytes and spermatocytes.
2. when the total sc per cell is compared, domesticus males have significantly lower SC arae compared to molossinus and musculus strains.Except in Q2 I Say this isn't true.>

@@

Adjusting for XX

Summarize for now, refine this section later.

1. illustrate problem(affects mostly SC length)  
2. Expected impact on sex comparisons, estimated effect size of the X  
3. (prove general pattern that ALL bivalents are longer), chrms sorted by bin comparisons
4. permutations of 19 female, 19male, 20female 20 male

<Of all female single bivalents observations, 5% are XX (1 of 20). 
- The XX is large, likely within the top 25% longest bivalents of the cell (3rd largest by Mb).
- The average % of XX for whole cell SC (sum(all bivalents)) can be calculated from the whole.cell data set. Lets guess 12% of a cell's total SC area is XX.
- rate of bivalent segmentation /* rate of XX, 5% /* mean SC length for 3rd longest bivalent / total SC area (by bivalent) = proportion of SC area due to XX>

@@

<!-- motivating

To quantify one of the major -- patterns that is different between male and female recombiantion landscapes
In order to test for evolution -- and examine sex differences in the recombination landscape (crossover position bias)
-->

@@

In the domesticus strains the mean normalized foci positions were significantly different between sexes and fit the predictions of the 'typical recombination landscape' (with female positions being medially placed and terminally placed in males). The p.values across the strains are; (t.test; p = `r Q1_1CO.pos_t.test.WSB$p.value`, `r Q1_1CO.pos_t.test.G$p.value`, and `r Q1_1CO.pos_t.test.LEW$p.value` for WSB, G and LEW respectively).

With the musculus and molossinus strains displaying a similar pattern (t.test p = `r Q1_1CO.pos_t.test.PWD$p.value` and `r Q1_1CO.pos_t.test.MSM$p.value`) PWD and MSM. Two musculus  strains had a weaker signal, but it was still in the same direction (t.test; p =  `r Q1_1CO.pos_t.test.KAZ$p.value`, and `r Q1_1CO.pos_t.test.SKIVE$p.value` for KAZ and SKIVE respectively).

After confirming the significan sex differences for 1CO position, we apply the model frameworks

@@

@@

Crossover interference is one of the major determinates of the recombination landscape, -- shifting it away from random / uniform

shifting the CO positions -- away from a random (distribution)

. We isolated `r length(  (Curated_BivData.22 %>% filter(sex == "female") )$Obj.ID )` and 
`r length(  (Curated_BivData.22 %>% filter(sex == "male") )$Obj.ID )` 2CO bivalents in females and males to test discribe sex differences in crossover interference. Mean interfocal distances (IFD) were calculated from `r length( unique( (Curated_BivData.22 %>% filter(sex == "female") )$mouse) )`  female and `r length( unique( (Curated_BivData.22 %>% filter(sex == "male") )$mouse) )` male mice. We examined both raw IFD^raw^ and normalized normalized by SC length (IFD^PER^). The mean IFD^raw^ measures of crossover interference as a mechnical force mediated through the SC. We compared the IFD^raw^ to test the hypothesis of interference acting as a mechanical force transmitted through the SC. Previous measures found no sex difference in measures of IFD^raw^ in mice (de Boer et al ).

@@

We examined the 2CO observations  // the distributions of 2CO IFD^PER^ to better discribe how the means / distributions of this metric are different between the sexes.

We note that in females there are/ may be signals of upper and lower thresdholds for IFD^PER^.

@@

The differences between the sexes is slightly different depending on if the males are in the high rec group. In the low rec strains (Dom + KAZ)...
(both males and females IFD distributions have lower and upper thresholds around 25% and 75%) respectively. however the weight of observations are slightly shifted across the sexes. There's an enrichment of lower IFD^PER^ in females relative to males, while the opposite is true for the higher threshold -- where males have an enrichment of higher/long IFD^PER^ compared to females.

@@

<!--
G - the upper and lower thresholds of females look pretty similar, the lower is closer to .2 than .25 -- female has more lower IFDs

@@

@@

glm M2, 
sex (glm; p= `r summry.glm.Q1_IFD.PER.M2$coefficients[,4][4]`) p = .0080 (this is different in the knited version!! arg)

male*PWD p = 0.02 (glm; p =`r summry.glm.Q1_IFD.PER.M2$coefficients[,4][14]`)

male * SKIVE p = 0.0058  (glm; p=`r summry.glm.Q1_IFD.PER.M2$coefficients[,4][15]`)

for **glm M3**

sex / male p = `r summry.glm.Q1_IFD.PER.M3$coefficients[,4][2]` and male*SKIVE just slightly significant (p =  `r summry.glm.Q1_IFD.PER.M3$coefficients[,4][14]`

@@

For the normalized values in both M2 and M3, sex is a significant effect, increasing nrm.IFD in males. SKIVE*male is the only other consistently significant effect, which also increases the nrm.IFD measure.

@@

@@

We tested the sex differences -- now we focus on the largest aspect of variation, the high rec males vs the low rec males.

@@

<put in plot for total SC and mean gwRR count>

<motivation: the positive correlation, Our prediction for this metric is that the high rec strain will have longer sc lengths (axis lengths)>

@@

while testing the sex differences, there were results (the significant subspecies effect) which suggest that the musculus and molossinus strains had longer SC compared to the domesticus strains - these two strains also contain the high rec strains - motivating a model/hypothesis that sc length / chromatin compaction evolution may proceed - facilitate the rapid male specific evolution in gwRR

<!-- basic stats
`r length(OG.MLH1.merge.Skel$fileName)` total skelentonize measures were made from all of the MLH1 pachtyene cell images. X observations removed for having extra signal from other cells. `r length(unique( all.remove ) )` measures removed as potential outliers (by examining the category specific plots)

 place set ups for total SC, short biv and long biv here -->

@@

```{r Q2.SC_MLH1.by.totalSC, echo=FALSE}

@@

**mouse means of total SC area for Higher rec strains are significantly higher compared to low rec strains (ttest; p = 
`r Q2tot.sc_t.test_High.Low$p.value`).**

Molossinus have significant difference between mouse means for total sc, between the high and low strains (mol p = `r Q2tot.sc_t.test_High.Low_Mol$p.value`). The total sc mouse means are not significant for the high and low strains in musc (musc p= `r Q2tot.sc_t.test_High.Low_Musc$p.value`).

@@

<across subspecies: only significant at the bivalent level, (Dom and Musc are sig different, Dom - Mol are sig difference) (also when high rec and low rec are compared).>

@@

@@

**For the total SC glm**

<Thse are our prediction for the M3 glm (just straint) models based on the variation of the gwRR's
1. no difference in DOM strains (for SC lengths )
2. PWD > SKIVE > the rest
3. Msm > mol>

@@

<And the strain effect for G is significant (glm; G p =`r sum.glmQ2_short.biv_M1$coefficients[,4][4]`).>

<For the M2 model analyzing the mouse means short bivalents, all strains have significant effect but LEW, G,PWD, MSM, MOLF, SKIVE, KAZ, CZECH.>

<!--24sep14_PWD_m1 is lower than the others, 5jul17_MSM_m1 is the low MSM-->

<just like the sex differences in SC area -- this points to differential usage of the SC for placing CO -- investigate the recombination landscape>

<!--
blocks of other male Q2 - SC lengths
what are the pairwise t-test functions -- use pairwise.t.test(Ozone, Month, p.adj = "bonf")
The mouse average and single bivalents are significantly longer in the high rec group p = r T.test_SC.Len_mouse.av$p.value and p=  r T.test_SC.Len_single.biv$p.value.
-->

**The mouse means for the reduced bivalent datasets (short and long) are not significantly different between the high and low rec strains (for the total pooled data) (ttest; short p = `r Q2short.biv_t.test_High.Low$p.value` and long p = `r Q2long.biv_t.test_High.Low$p.value`).**

@@

- confirm that the positive correlation of SC and CO number is maintained

@@

@@

# brief transition, rec landscape

the lack of predictive power of SC length to predict CO number indicates that there might be different patterns of 'SC usage' / different patterns in the layout of the recombination landscape /  these patterns could stem from placement or spacing of foci.

@@

<!-- motivation / background of the sex differences, males more terminall in CO placement compared to females, now we ask if there are significant differences in the placement of single foci along bivalents

Main results:  
1. WSB has the most terminal 1CO rec landscape and MOLF has the most central.
2. Evidence for evolution of the 1CO rec landscape, but it is decoupled from the evolution in gwRR we observe in figure 1
2.5  The normalized 1CO position is not a good metric for distinguishing strains with rapid gwRR evolution

-->
<!-- set up chunks here -->
<!-- plots /figures below-->

@@

The previous section found that the single crossover landscape is significantly different between males and females. In this section we focus on differences between the high and the low rec male strains
In order to test if if there are distinguishing meterics for these two groups. **The figure above shows, WSB has significantly more terminal positioning, while MOLF has significantly more medial position.**

T tests of the single crossover positions reveal that the normalized single foci positions are not significantly different between the total pooled high and low rec male strains (t.test; p = `r Rec.group_BivData_F1post_Ttest$p.value`). The high rec musc strains, _musculus^PWD^_ and _musculus^SKVIE^_, do not have signifcantly different single foci landscapes from the other musc low rec strains (ttest; p = `r Q2.1COpos_mus.av_Musc_ttest$p.value`). While _molossinus^MOLF^_ is significantly more centrally positioned compared to _molossinus^MSM^_ (t.test; p = `r Q2.1COpos_mus.av_Mol_ttest$p.value`). This result is somewhat puzzeling and it doesn't match up with the strains with high recombination rates.

<long p = `r Rec.group_LONG.BivData_F1post_Ttest$p.value` short 1CO, male, p = `r Q2.1COpos.PER_short_ttest_HL$p.value`> 
<!-- GLMs because there's no difference 
toal data set, short and long biv. There's only a difference between MSM and MOKF
test ppredictions, Dom = NULL,  -->

@@

<!--
In the above chunk I tried making another glm's which uses the single bivalent measures and include chromosome size for the switching the model to include interaction (instead of additive), all of the pvalues get lower. .. the chromosome length has the most significant effect, LEW by chrm effect are significant

@@

LEW (slight) (p =`r summary(Q2.glm_mus.av_1COpos_M1)$coefficients[,4][5]`).  
But big effects from Mol (p = `r summary(Q2.glm_mus.av_1COpos_M1)$coefficients[,4][3]`).  
and MSM strain (p = `r summary(Q2.glm_mus.av_1COpos_M1)$coefficients[,4][7]`)  
**For M3 (all strains)**

strain G (p = `r sum.Q2.glm_mus.av_1COpos_M2[,4][2]`).

strain Lew (p = `r sum.Q2.glm_mus.av_1COpos_M2[,4][3]`).

strain PWD ( p = `r sum.Q2.glm_mus.av_1COpos_M2[,4][4]`).

slight strain MSM  (p = `r sum.Q2.glm_mus.av_1COpos_M2[,4][5]`).

very sig strain MOLF (p = `r sum.Q2.glm_mus.av_1COpos_M2[,4][6]`).

strain SKIVE (p = `r sum.Q2.glm_mus.av_1COpos_M2[,4][7]`).

strain KAZ (p = `r sum.Q2.glm_mus.av_1COpos_M2[,4][8]`).
-->

@@

given that the t.test don't return significant pvalues so we are not motivated to follow up with linear models. Males from WSB and MOLF strains have 'strang'unique patterns for the single foci positions, which don't fit with predicted patterns from the genome wide recombination rate variation

<I ran another model that used the single bivalents observations to predict the normalize foci position with fixed effects of chromosome length, subspecies, strian and the interaction effects.>

<!--
**M.X. Linear Model** $$1CO \ Normalized \ Position ~=~ chromosomeLength * subsp * strain + \varepsilon $$ -->

@@

subsp MOL `r summary(chrm.1CO.pos.chrmLength)$coefficients[,4][4]`  
the mol obj above isn't showing the reight value.

strain G   `r summary(chrm.1CO.pos.chrmLength)$coefficients[,4][5]`

strain LEW   `r summary(chrm.1CO.pos.chrmLength)$coefficients[,4][6]`

Strain MSM `r summary(chrm.1CO.pos.chrmLength)$coefficients[,4][8]`

chrm * LEW `r summary(chrm.1CO.pos.chrmLength)$coefficients[,4][14]`

chrm *MSM  `r summary(chrm.1CO.pos.chrmLength)$coefficients[,4][16]`

M2

-chromosome length, `r summary(chrm.1CO.pos.chrmLength_strain)$coefficients[,4][2]`

-strain G, `r summary(chrm.1CO.pos.chrmLength_strain)$coefficients[,4][3]`

-strain LEW, `r summary(chrm.1CO.pos.chrmLength_strain)$coefficients[,4][4]`

-strain MOLF, `r summary(chrm.1CO.pos.chrmLength_strain)$coefficients[,4][7]`

-strain SKIVE, `r summary(chrm.1CO.pos.chrmLength_strain)$coefficients[,4][8]`

-chrm length LEW, `r summary(chrm.1CO.pos.chrmLength_strain)$coefficients[,4][12]` -->

@@

<!-- set up chunks here-->

@@

biv.IFDs_PER_by.group

@@

```{r Q2.t.test_IFD,echo=FALSE, eval=FALSE}

@@

<write some kind of intro here>motivation /

A common/logical prediction for the relationship of crossover number and interference strength is -- a negeative correlation. ()

how crossover interference affects the evolution of genome wide recombination rate -- is

prediction for the interference pattern in the high rec strains is that the high rec strains will have weaker interference (ie. less space between foci on the same bivalent, which allows more foci to be placed. GIVEN THAT THERE ARE strain effects for the SC length – we only examine the IFD^PER^ which controls for variation in SC lengths.

The mouse averages for IFDs were significantly longer in the high rec groups (t.test; p = `r t.test_mouse.av.IFD_PER$p.value` for IFD^PER^ and p = `r t.test_mouse.av.IFD_ABS$p.value` for IFD^raw^). The pattern is confirmed with t.test comparing groups within the same subepcies .. add these ttests and p values.

<glms> The following model framework was applied to further investigate these patterns using mouse averages of IFD^raw^ and IFD^PER^.

The mouse averages for IFDs were significantly longer in the high rec groups (t.test; p = `r t.test_mouse.av.IFD_PER$p.value` for IFD^PER^ and p = `r t.test_mouse.av.IFD_ABS$p.value` for IFD^raw^). The pattern is confirmed with t.test comparing groups within the same subepcies.

<glms> The following model framework was applied to further investigate these patterns using mouse averages of IFD^raw^ and IFD^PER^.

@@

#Mol
Q2.glm.M2_IFD.PER_Mol <- glm(mean_IFD.2CO_PER ~ strain, data= Male.poly.Mouse.Table_BivData_4MM[Male.poly.Mouse.Table_BivData_4MM$subsp == "Mol",])
#summary(Q2.glm.M2_IFD.PER_Mol)
```

In M1, the model build on the mouse averages of IFD metrics with subspecies and strain as fixed effects, the high rec strains had significant strain effects (glm; p = `r sum.Q2.M1glm_IFD.PER$coefficients[,4][6]`, p = `r sum.Q2.M1glm_IFD.PER$coefficients[,4][7]` and p = `r sum.Q2.M1glm_IFD.PER$coefficients[,4][8]`  for PWD, MSM and SKIVE respectively). In M2, the mouse average for normalzied IFD is modeled with only strain effects. Only PWD and SKIVE IFD^PER^ had significant strain effects (glm; p = `r sum.Q2.M2glm_IFD.PER$coefficients[,4][4]` and p = `r sum.Q2.M2glm_IFD.PER$coefficients[,4][7]` for PWD and SKIVE respectively).

<!--when the M2 model was run within subspecies, the results generally confirmed our predictions; no sig values for Dom, PWD and SKIVE have longer IFDs than other musc strains and MSM is longer than MOLF-->

<!-- we ran similar models using 2CO bivalent observations to examine what effects the SC length might have on IFD^PER^ length.

**model BivLen.M1 ** $$Single bivalent\ IFD ~=~ subsp * strain * chromosomeLength + \varepsilon $$

**model BivLen.M2 ** $$Single bivalent\ IFD ~=~ strain * chromosomeLength + \varepsilon $$

```{r Q2.IFD.glm.chrmLength, echo=FALSE}

>>>>

In order to deconstruct the cell wide MLH1 counts we look at the recombination landscape at the single bivalent level. From our total set of cell images

<FROM number of crossovers per chromsoomes TO rec lanscape across single chromsomes>
<A big motivation was to quantifiy recombiantion landscape differences across single bivalents>

@@

In our analysis our first step is to test for measurable differences, between sexes or rec.groups and is the differences are significant, our second step is to apply the same sets of models applied to the mean MLH1 foci counts to test / describe the effects due to subspecies or strains.

@@

An additional measure of single bivalents, we tested a measure of the SC lengths for the whole genome (whole cells) using an image analysis algorithm which measures the skeletonized bivalent and provides an acurate measure of the lengths of single SC/bivalents (ref RWang).

We apply a skeletonizing image analysis algorithm to all MLH1 meiocyte spreads images. X cell wide total SC measures remained after and after outliers removed,
Mouse means were calculated for their total SC area per cell and are displayed in figure X.

@@

@@

In the mixed models for the sex differences mean **short bivalents** length as the dependaent variable, sex is the most significant effect (LTR; p =`r Q1lmer.MM_SC_short_results$sex.results`). The interaction (subspecies by sex) effect was also slightly significant (p = `r Q1lmer.MM_SC_short_results$interaction.pval`).

In the mixed model run on the mean total SC metric, there are more significant effects for strain and subsp for the mixed model using mean **total sc** all  of the coefficients were significant, (LTR; subsp p= `r Q1.lmer.totSC_results$sub.sp`, sex p = `r Q1.lmer.totSC_results$sex.results`, and interaction p = `r Q1.lmer.totSC_results$interaction.pval`, (random LRT; strain p = `r Q1.lmer.totSC_results$rand.pval`).

@@

In linear models (M2 and M3) for **mean short bivalent length**, sex is the most significant coefficient (M2 glm; p =`r summ.Q1.glm_short.SC_M2$coefficients[,4][8]` and M3 glm; p = `r summ.Q1.glm_short.SC_M3$coefficients[,4][8]`) with slight significant strain and interaction effects (glm; p=`r summ.Q1.glm_short.SC_M2$coefficients[,4][7]`, p =`r summ.Q1.glm_short.SC_M2$coefficients[,4][14]` respectively).

@@

@@

@@

<!-- motivating
To quantify one of the major -- patterns that is different between male and female recombiantion landscapes
-->

@@

In all strains, the 1CO landscape had significant differences between males and female which followed the ‘typical landscape’ of females having central positions and males have terminal foci positions.
(These results were also supported by the Mixed model which 
For the mixed model, sex is the most significant effect (LTR; p =1.2610^{-25}).The random effect of strain is also significant (LRT; p =0.01).

After confirming the significant sex differences for 1CO position, we apply the model frameworks

@@

The linear models were used --- to detect if specific strains differed in the sex differences in the 1CO landscape. 
While sex was the most significant effect for both M2 and M3

@@

Crossover interference is one of the major determinates of the recombination landscape. It generates a distribution of evenly spaced crossover along chromosomes instead of a random and more uniform distribution.

 We isolated `r length(  (Curated_BivData.22 %>% filter(sex == "female") )$Obj.ID )` and 
`r length(  (Curated_BivData.22 %>% filter(sex == "male") )$Obj.ID )` 2CO bivalents in females and males to test describe sex differences in crossover interference. Mean interfocal distances (IFD) were calculated from `r length( unique( (Curated_BivData.22 %>% filter(sex == "female") )$mouse) )`  female and `r length( unique( (Curated_BivData.22 %>% filter(sex == "male") )$mouse) )` male mice. We examined both raw IFD^raw^ and normalized normalized by SC length (IFD^PER^). The mean IFD^raw^ measures of crossover interference as a mechanical force mediated through the SC. We compared the IFD^raw^ to test the hypothesis of interference acting as a mechanical force transmitted through the SC. Previous measures found no sex difference in measures of IFD^raw^ in mice (de Boer et al ).

@@

GENERAL RESTULTS – DIFFERENCES between sexes

In examining the IFD^PER^ distributions, we note that 
the female norm IFD are centered at ~50% – but they also seem to have slight an enrichment of short (<25%) normalized IFDs

While the general pattern in the male distributions are enrichment longer IFD^PER^. (however we note that there are slight differences in the IFD^PER^ distributions between the high and low rec males to be followed up in later section)

@@

<!--
The differences between the sexes is slightly different depending on if the males are in the high rec group. In the low rec strains (Dom + KAZ)...
(both males and females IFD distributions have lower and upper thresholds around 25% and 75%) respectively. however the weight of observations are slightly shifted across the sexes. There's an enrichment of lower IFD^PER^ in females relative to males, while the opposite is true for the higher threshold -- where males have an enrichment of higher/long IFD^PER^ compared to females.

@@

G - the upper and lower thresholds of females look pretty similar, the lower is closer to .2 than .25 -- female has more lower IFDs

@@

To test for strain specific effects we applied the model framework

@@

@@

For the normalized values in both M2 and M3, sex is a significant effect, increasing nrm.IFD in males. SKIVE*male is the only other consistently significant effect, which also increases the nrm.IFD measure (p = male-SKIVE p =  )

@@

Our comparison of interference across sexes – supports / applies with general patterns /results from the literature:
When interference is measured in physical SC units,  – the differences between sexes – is low / only slightly significant 
 IFD raw is not significantly different (t.tests and models?). – indicating that the 
While when the variance in chromatin compaction is controlled – males have stronger interference – (like genetic maps) – the impact on the rec landscapes being ….

@@

The results from the previous section -- confirm (known sex differences for the 'typical' 
recombination landscape (Sardell Kirkpatrick). The next section is meant to focus on the greater aspect of variation in mean MLH1 counts per cell the high rec males males strains ( _musculus^PWD^_ , _musculus^SKIVE^_, and _molossinus^MSM^_) and the rest of the males with the main objective of this section is to test for significant correlations with the evolution of mean MLH1 foci per cell.

In comparing the male specific single bivalent based metrics the first task/level is to tests for significant differences between the high and low rec groups. The second being to follow up with linear models that test for effects of subspecies and strains.

The literature provides basic predictions for the direction of the relationship the genome wide recombination rate and two aspects of the recombiantion landscape; SC length (positve) and crossover interference (negative). Following this logic we predict i) no strain effects for SC length or crossover interference within _domesticus_, ii) _musculus^PWD^_ will have greater SC length and weaker interference than _musculus^SKIVE^_, which will have longer SC and weaker interference compared to the other _musculus_ strains and iii) _molossinus^MSM^_ will have longer SC and weaker interference compared to _molossinus^MOLF^_.

@@

@@

While testing the sex differences, there were results (the significant subspecies effect) which suggest that the musculus and molossinus strains had longer SC compared to the domesticus strains - these two strains also contain the high rec strains - motivating a model/hypothesis that sc length / chromatin compaction evolution may proceed - facilitate the rapid male specific evolution in gwRR

The confounding effects of sex chromosomes are removed in this section that makes comparisons within males. Three sc based metrics; total SC, mean short bivalent length and mean long bivalent length. With long bivalents isolated in the same manner as the short bivalent.

<!-- basic stats
`r length(OG.MLH1.merge.Skel$fileName)` total skelentonize measures were made from all of the MLH1 pachtyene cell images. X observations removed for having extra signal from other cells. `r length(unique( all.remove ) )` measures removed as potential outliers (by examining the category specific plots)-->

@@

```{r Q2.SC_MLH1.by.totalSC, echo=FALSE, warning=FALSE}

@@

There is a generally positive correlation between mean MLH1 foci per cell and total SC, however the variation in mean total SC does not completely predict the high and low rec strains (Figure X).

Within the total data set of mean total SC, high rec strains have significantly more total SC area (ttest; p = `r Q2tot.sc_t.test_High.Low$p.value`). However when the high and low rec strains are compared within subspecies only the molossinus strains have significant difference  (ttest mol p = `r Q2tot.sc_t.test_High.Low_Mol$p.value`), while the musculus strains do not have significant differences (ttest; musc p= `r Q2tot.sc_t.test_High.Low_Musc$p.value`).

The mouse means for the reduced bivalent datasets (short and long) are not significantly different between the high and low rec strains (for the total pooled data) (ttest; short p = `r Q2short.biv_t.test_High.Low$p.value` and long p = `r Q2long.biv_t.test_High.Low$p.value`).

@@

@@

becasue there is weak // some significant differences between the sc metrics of the high and the low rec groups --- we proceed with running the model framework.

**Linear model M1 ** $$mouse \ average \ SC metric ~=~ subsp * strain + \varepsilon $$

**Linear model M2 ** $$mouse \ average \ SC metric ~=~ strain + \varepsilon $$

@@

@@

<And the strain effect for G is significant (glm; G p =`r sum.glmQ2_short.biv_M1$coefficients[,4][4]`).> --

M3 -- tests within subspecies!!

@@

@@

**THERE IS a weak signal of the high rec strains having longer SC (having evolved) longer SC (different chromatin compaction) –  While the mouse mean total sc are significantly different between the two groups --- this difference isn’t translated to the reduced bivalent data sets (p = long  p = short)**

@@

@@

@@

The normalized 1CO position is not significantly different between the high and low rec groups for the total pooled data (t.test; p = `r Rec.group_BivData_F1post_Ttest$p.value`) and also when examined within subspecies (ttest; p = `r Q2.1COpos_mus.av_Musc_ttest$p.value` and p = `r Q2.1COpos_mus.av_Mol_ttest$p.value` for _musculus_ and _molossinus_ respectively). While there were significant strain effects for _domesticus^^WSB^_ and _molossinus^MOLF^_ (figure X), this evolution of the 1CO positioning is decoupled from the total genome wide recombination rate.

<!--
The high rec musc strains, _musculus^PWD^_ and _musculus^SKIVE^_, do not have signifcantly different single foci landscapes from the other musc low rec strains (ttest; p = `r Q2.1COpos_mus.av_Musc_ttest$p.value`). While _molossinus^MOLF^_ is significantly more centrally positioned compared to _molossinus^MSM^_ (t.test; p = `r Q2.1COpos_mus.av_Mol_ttest$p.value`). 
is that _domesticus^WSB^_ has significantly more terminal positioning,
and _molossinus^MOLF^_ having significantly more medial position.
A general pattern of strain specific effects
we found in looking at the 1CO positions, is that _domesticus^WSB^_ has significantly more terminal positioning, while MOLF has significantly more medial position.-->

<!-- don't report GLMs - since not difference in the ttests, GLMs because there's no difference. long p = `r Rec.group_LONG.BivData_F1post_Ttest$p.value` short 1CO, male, p = `r Q2.1COpos.PER_short_ttest_HL$p.value`> 
toal data set, short and long biv. There's only a difference between MSM and MOKF
test ppredictions, Dom = NULL,  -->

@@

<!--
given that the t.test don't return significant pvalues so we are not motivated to follow up with linear models. Males from WSB and MOLF strains have 'strang'unique patterns for the single foci positions, which don't fit with predicted patterns from the genome wide recombination rate variation
<I ran another model that used the single bivalents observations to predict the normalize foci position with fixed effects of chromosome length, subspecies, strian and the interaction effects.>

In the above chunk I tried making another glm's which uses the single bivalent measures and include chromosome size for the switching the model to include interaction (instead of additive), all of the pvalues get lower. .. the chromosome length has the most significant effect, LEW by chrm effect are significant

@@

LEW (slight) (p =`r summary(Q2.glm_mus.av_1COpos_M1)$coefficients[,4][5]`).  
But big effects from Mol (p = `r summary(Q2.glm_mus.av_1COpos_M1)$coefficients[,4][3]`).  
and MSM strain (p = `r summary(Q2.glm_mus.av_1COpos_M1)$coefficients[,4][7]`)  
**For M3 (all strains)**
strain G (p = `r sum.Q2.glm_mus.av_1COpos_M2[,4][2]`).
strain Lew (p = `r sum.Q2.glm_mus.av_1COpos_M2[,4][3]`).
strain PWD ( p = `r sum.Q2.glm_mus.av_1COpos_M2[,4][4]`).
slight strain MSM  (p = `r sum.Q2.glm_mus.av_1COpos_M2[,4][5]`).
very sig strain MOLF (p = `r sum.Q2.glm_mus.av_1COpos_M2[,4][6]`).
strain SKIVE (p = `r sum.Q2.glm_mus.av_1COpos_M2[,4][7]`).
strain KAZ (p = `r sum.Q2.glm_mus.av_1COpos_M2[,4][8]`).
-->

@@

@@

subsp MOL `r summary(chrm.1CO.pos.chrmLength)$coefficients[,4][4]`  
the mol obj above isn't showing the reight value. 
strain G   `r summary(chrm.1CO.pos.chrmLength)$coefficients[,4][5]`  
strain LEW   `r summary(chrm.1CO.pos.chrmLength)$coefficients[,4][6]`  
Strain MSM `r summary(chrm.1CO.pos.chrmLength)$coefficients[,4][8]`  
chrm * LEW `r summary(chrm.1CO.pos.chrmLength)$coefficients[,4][14]`  
chrm *MSM  `r summary(chrm.1CO.pos.chrmLength)$coefficients[,4][16]`

M2
-chromosome length, `r summary(chrm.1CO.pos.chrmLength_strain)$coefficients[,4][2]`  
-strain G, `r summary(chrm.1CO.pos.chrmLength_strain)$coefficients[,4][3]`  
-strain LEW, `r summary(chrm.1CO.pos.chrmLength_strain)$coefficients[,4][4]`  
-strain MOLF, `r summary(chrm.1CO.pos.chrmLength_strain)$coefficients[,4][7]`  
-strain SKIVE, `r summary(chrm.1CO.pos.chrmLength_strain)$coefficients[,4][8]`  
-chrm length LEW, `r summary(chrm.1CO.pos.chrmLength_strain)$coefficients[,4][12]` -->

@@

@@

@@

t.test_mouse.av.IFD_PER_DOM <- t.test(  (mouse.avs_4MM %>% filter(Rec.group == "1") %>% filter(sex == "male") %>% filter(subsp == "Dom") )$mean_IFD.2CO_PER,
   (mouse.avs_4MM %>% filter(Rec.group == "0") %>% filter(sex == "male")  %>% filter(subsp == "Dom"))$mean_IFD.2CO_PER   )

t.test_mouse.av.IFD_PER_MUSC <- t.test(  (mouse.avs_4MM %>% filter(Rec.group == "1") %>% filter(sex == "male") %>% filter(subsp == "Musc") )$mean_IFD.2CO_PER,
   (mouse.avs_4MM %>% filter(Rec.group == "0") %>% filter(sex == "male")  %>% filter(subsp == "Musc"))$mean_IFD.2CO_PER   )

t.test_mouse.av.IFD_PER_MOL <- t.test(  (mouse.avs_4MM %>% filter(Rec.group == "1") %>% filter(sex == "male") %>% filter(subsp == "Mol") )$mean_IFD.2CO_PER,
   (mouse.avs_4MM %>% filter(Rec.group == "0") %>% filter(sex == "male") %>% filter(subsp == "Mol") )$mean_IFD.2CO_PER   )

```{r Q2.t.test_IFD, echo=FALSE, eval=FALSE}

@@

<write some kind of A common/logical prediction for the relationship of crossover number and interference strength is -- a negeative correlation. how crossover interference affects the evolution of genome wide recombination rate -- is prediction for the interference pattern in the high rec strains is that the high rec strains will have weaker interference (ie. less space between foci on the same bivalent, which allows more foci to be placed. GIVEN THAT THERE ARE strain effects for the SC length – we only examine the IFD^PER^ which controls for variation in SC lengths.>

Unlike total SC length and 1CO pattern, IFD metrics are significant predictors of the rapid genome wide recombination rate evolution. However, instead of a negative correlation between interference strength and number genome wide recombiantion rate (higher gwRR associated with weaker interference, interference strength is stronger in the high rec strains.

Unexpectedly, the mouse averages for both IFD^raw^ and IFD^PER^ were significantly longer in the high rec groups (t.test; p = `r t.test_mouse.av.IFD_PER$p.value` for IFD^PER^ and p = `r t.test_mouse.av.IFD_ABS$p.value` for IFD^raw^). The pattern is confirmed with t.test comparing groups within _musculus_ and _molossinus_
the same subepcies (add these ttests and p values.)
(t.tests p =`r t.test_mouse.av.IFD_PER_MUSC$p.value` and p= `r t.test_mouse.av.IFD_PER_MOL$p.value`).

We test the within subspecies predictions using the model framework

@@

#Mol
Q2.glm.M2_IFD.PER_Mol <- glm(mean_IFD.2CO_PER ~ strain, data= Male.poly.Mouse.Table_BivData_4MM[Male.poly.Mouse.Table_BivData_4MM$subsp == "Mol",])
#summary(Q2.glm.M2_IFD.PER_Mol)

In M1, the model build on the mouse averages of IFD^PER^ metrics with subspecies and strain as fixed effects, the high rec strains had significant strain effects for (glm; p = `r sum.Q2.M1glm_IFD.PER$coefficients[,4][6]`, p = `r sum.Q2.M1glm_IFD.PER$coefficients[,4][7]` and p = `r sum.Q2.M1glm_IFD.PER$coefficients[,4][8]` for _musculus^PWD^_, _molossinus^MSM^_ and _musculus^SKIVE^_, respectively).

In the M2 model results, _musculus^PWD^_, and _musculus^SKIVE^_, were the only significant strain effects, (glm; p = `r sum.Q2.M2glm_IFD.PER$coefficients[,4][4]` and p = `r sum.Q2.M2glm_IFD.PER$coefficients[,4][7]`).

Running M2 within subspecies – confirms the within subepcies predictions across strains – (no significant strain effects

Q2.glm.M2_IFD.PER_Dom

Q2.glm.M2_IFD.PER_Musc

Q2.glm.M2_IFD.PER_Mol

```{r Q2.IFD.glm.chrmLength, echo=FALSE, eval=FALSE}
